Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-09-05 marginal.Rmd
Small change
Some cutting in the first pages

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevail in old age---the retired population---as in working age.  Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
strong determinant when it comes to health and mortality and its impact are even
increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserve their appellation of
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98]. The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies. This is a
reasonable assumption since these societies were often characterized by
enormous class differences. Knowing that access to different kinds of resources, such as economic, social and cultural capital, provides advantages in all aspects of life, the health advantage of higher classes ought to be obvious.

@@

Social position affects health and mortality differently during the life course. There are different hypotheses about the impact in adulthood and old age respectively. The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral, rewarding some while disfavouring others. This leads to larger differences in old age.

Another aspect of the development of social inequality in mortality concerns its relation to economic inequality. @wipi09 argue that income inequality has an independent effect on mortality, separate from the direct effect of actual access to economic resources. They find
that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04; @wado00]. In recent decades, the respective association between trends of inequalities in mortality and income is weak or non-existent according to @hoetal16. What the association looked like in previous periods is unknown;  this is of particular interest as the levels of economic inequality differed fundamentally from those of recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were the most disadvantaged due to large levels of inequality.

When it comes to Sweden both income and wealth distributions were strongly skewed from the 1870s to the early 20th century. Starting during the inter-war period, income and
wealth inequality continuously diminished to reach a low point at around
1980. Since then, economic inequality has increased substantially until the
present, although it is far from as high as it was a century ago [@rowa08; @rowa09].
The Swedish development resembles that of other European and North
American countries in its basic features [@rowa14].

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. The regions studied here were for a long time remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships of the year came to this area in May or even as late as June in some cases [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

In our dataset, before 1950 the Skellefteå region consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century. The data from the period after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
for the earlier period but with the addition of two more parishes. The
majority of the 19th century population lived in rural villages and
hamlets, making its livelihood from agricultural production. During the
20th century, industrialization took place. This also led to a population increase both in both the town
and the rural parts, resulting in a more diversified economy.

Mortality was fairly low compared with other parts of the country and
the fertility transition was late [@coa86]. The Skellefteå population
size as defined in our data sets (all ages) was 6,142 on January 1, 1801, 43,212 on January 1, 1901, and 76,723 at the end of the 20th century.

The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, with another three parishes included. This region has a somewhat different character to that of Skellefteå. Umeå town had for a long time a small population, and was the administrative, educational and military centre of the county of Västerbotten. During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase. Agriculture dominated the rural part, but there were also some foundries as well as forestry industries. Consequently the economy was more diversified than that of Skellefteå. The population size as defined in our data sets (all ages) was 19,138 on January 1, 1901 and 103,970 when the 20th century ended.

@@

An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality. Even though our study does not directly address the hypothesis posed by @wipi09 and thus cannot falsify it, our results are relevant for this question as well. In the early 20th century, like other countries, Sweden was characterised by enormous economic differences; still, in our regions this did not lead to any consistent advantage for the highest social class. After the Second World War, economic differences began diminishing, and were at their lowest in Sweden around 1980; i.e. at a time when the expected pattern of a social hierarchy in survival appeared. The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences. We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality. The long-term relationship between economic inequality and health is still a puzzle requiring further studies.

>>>>

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population. The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is analysed according to gender and age group (working age vs retired). Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
strong determinant when it comes to health and mortality and its impact are even
increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserve their appellation of
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98]. The idea that socio-economic health inequalities were probably larger in historical societies is a
reasonable assumption since these societies were often characterized by
enormous class differences. Knowing that access to different kinds of resources, such as economic, social and cultural capital, provides advantages in all aspects of life, the health advantage of higher classes ought to be obvious.

@@

Social position affects health and mortality differently during the life course and from adulthood into old age. The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral, rewarding some while disfavouring others.

Another aspect of the development of social inequality in mortality concerns its relation to economic inequality. @wipi09 argue that income inequality has an independent effect on mortality, separate from the direct effect of actual access to economic resources. They find
that unequal societies perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04; @wado00]. In recent decades, the respective association between trends of inequalities in mortality and income is weak or non-existent according to @hoetal16. What the association looked like in previous periods is unknown;  this is of particular interest as the levels of economic inequality differed fundamentally from those of recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were the most disadvantaged due to large levels of inequality.

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. The regions studied here were for a long time remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen, in some cases as late as June [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures.The regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

In our dataset, before 1950 the Skellefteå region consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century. The data from the period after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
for the earlier period but with the addition of two more parishes. The
majority of the 19th century population lived in rural villages and
hamlets, making its livelihood from agricultural production. During the
20th century, industrialization took place. This also led to a population increase both in the town
and the rural parts, resulting in a more diversified economy. The Skellefteå population
size as defined in our data sets (all ages) was 6,142 on January 1, 1801, 43,212 on January 1, 1901, and 76,723 at the end of the 20th century.

The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, with another three parishes included. Umeå town had for a long time a small population, and was the administrative, educational and military centre of the county of Västerbotten. During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase. Agriculture dominated the rural part. Consequently the economy was more diversified than that of Skellefteå. The population size as defined in our data sets (all ages) was 19,138 on January 1, 1901 and 103,970 when the 20th century ended.

@@

An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality. Even though our study does not directly address the hypothesis posed by @wipi09 and thus cannot falsify it, our results are relevant for this question as well. In the 19th and early 20th century, like other countries, Sweden was characterised by enormous economic differences; still, in our regions this did not lead to any consistent advantage for the highest social class. After the Second World War, economic differences began diminishing, and were at their lowest in Sweden around 1980; i.e. at a time when the expected pattern of a social hierarchy in survival appeared. The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences [@rowa08; @rowa09]. We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality. The long-term relationship between economic inequality and health is still a puzzle requiring further studies.
